---
title: "CatMapR End-to-End Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CatMapR End-to-End Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

run_live <- identical(tolower(Sys.getenv("CATMAPR_RUN_LIVE", "false")), "true")
```

## Overview

CatMapper is a platform for organizing and harmonizing category systems across datasets.
It currently includes two applications: **SocioMap**, which focuses on social-science
categories (for example ethnicities, languages, religions, and administrative units),
and **ArchaMap**, which focuses on archaeological categories (for example artifact
types and cultural periods).
CatMapR is an API wrapper for the CatMapper API, and all package functions are
also available through the CatMapper web application at <https://catmapper.org>.

This vignette shows a standard CatMapR workflow:

1. list available dataset metadata,
2. inspect metadata helpers,
3. search and inspect categories,
4. translate external labels,
5. run basic quality checks,
6. join harmonized tables.

By default, chunks run in offline demo mode for reproducible rendering.

## Why Use CatMapper?

Researchers often work with datasets that describe the same category systems
in different and incompatible ways. CatMapper helps by:

- identifying likely matches across naming variants and coding systems,
- preserving context needed for disambiguation (for example place, domain, or period),
- reducing manual harmonization effort before analysis,
- making joins across independently structured datasets more reproducible.

CatMapR provides these capabilities in scriptable R workflows, so translation,
quality checks, and joins can be rerun consistently as data updates.

## What CatMapR Returns

CatMapR wrappers return CatMapper API responses, including:

- dataset catalog metadata (for example CMID, CMName, citation fields, years),
- category/entity metadata and relationship details,
- translation and join outputs,
- metadata/property discovery tables.

CatMapR does **not** fetch raw dataset source files managed outside CatMapper.
User-owned/raw datasets are external inputs to your R workflow.

## UI-to-R Function Mapping

* In routes like `/:database/explore`, `:database` means the app path segment, `sociomap` or `archamap` (for example `/sociomap/explore`).

| CatMapperJS route | Typical UI step | CatMapR function(s) |
| --- | --- | --- |
| `/:database/explore` | Search categories/entities and inspect details | `searchDatabase()`, `CMIDinfo()`, `getDomains()` |
| `/:database/translate` | Upload labels, run matching, review translated outputs | `translate()` |
| `/:database/merge` | Propose merge keys and join aligned datasets | `createLinkfile()`, `joinDatasets()` |
| `/:database/edit` | Authenticated edit upload with automatic waiting-USES contextual refresh | `getUploadProperties()`, `uploadInputNodes()`, `submitEditUpload()` |

## UI Workflow Crosswalk (Detailed)

| User intent | CatMapper UI path | UI action | CatMapR call pattern |
| --- | --- | --- | --- |
| Browse available dataset records | `/:database` and `/:database/explore` | Open database and review dataset catalog entries | `listDatasetMetadata(database)` (legacy: `allDatasets(database)`) |
| Inspect one known dataset CMID | `/:database/:cmid` | Open dataset info page | `getDatasetMetadata(database, CMID, domain, children)` (legacy: `datasetInfo(...)`) |
| Browse metadata helpers | `/:database/explore` and `/:database/edit` | Inspect available domains and property fields | `getDomains(database)`, `getUploadProperties(database)` |
| Find entities/categories | `/:database/explore` | Run search filters and inspect hits | `searchDatabase(...)`, then `CMIDinfo(...)` |
| Harmonize labels | `/:database/translate` | Upload table and run translation | `translate(rows, database, term, ...)` |
| Build merge candidates | `/:database/merge` | Propose merge linkfile | `createLinkfile(categoryLabel, datasetChoices, ...)` |
| Join aligned sources | `/:database/merge` | Run join on matched keys | `joinDatasets(database, joinLeft, joinRight, ...)` |
| Submit metadata/category edits | `/:database/edit` | Upload edit rows and auto-trigger contextual USES refresh | `submitEditUpload(...)` (preferred) or `uploadInputNodes(...)` |

## Setup

```{r setup}
library(CatMapR)

# Optional API override:
# Sys.setenv(CATMAPR_API_URL = "https://api.catmapper.org")

run_live
```

## Candidate Dataset Pairings

Suggested pairings for a worked example:

- SocioMap ADM0: `SD1` (GADM 3.6) + `SD2` (GeoNames 202005)
- SocioMap ETHNICITY: `SD2176` (GeoEPR 2021) + `SD461930` (LEDA)
- SocioMap LANGUAGE: `SD3` (Glottolog 4.2.1) + `SD2196` (Wikidata)
- ArchaMap artifact/period: `AD37767` (Andrefsky 2005) + `AD37770` (DAI_FeatureTypes)

Dataset IDs above were confirmed on 2026-03-18 and may change as catalogs update.

## 1) List Datasets

```{r list-datasets}
if (run_live) {
  datasets <- listDatasetMetadata(database = "SocioMap")
} else {
  datasets <- data.frame(
    CMID = c("SD1", "SD2", "SD2176", "SD461930"),
    CMName = c("GADM 3.6", "GeoNames Version 202005", "GeoEPR 2021", "LEDA"),
    ApplicableYears = c("2018", "2020", "1946-2021", NA),
    stringsAsFactors = FALSE
  )
}

head(datasets)
```

## 2) Inspect Metadata Helpers

```{r metadata-helpers}
if (run_live) {
  domains <- getDomains(database = "SocioMap")
  upload_props <- getUploadProperties(database = "SocioMap")
} else {
  domains <- data.frame(
    domain = c("DISTRICT", "DISTRICT", "ETHNICITY"),
    subdomain = c("ADM0", "ADM1", "ETHNICITY"),
    description = c("Administrative district", "Administrative district", "Ethnicity category"),
    stringsAsFactors = FALSE
  )
  upload_props <- list(
    database = "SocioMap",
    nodeProperties = data.frame(property = "DatasetCitation", description = "Dataset citation", stringsAsFactors = FALSE),
    usesProperties = data.frame(property = "yearStart", description = "Starting date", stringsAsFactors = FALSE)
  )
}

head(domains)
head(upload_props$nodeProperties)
head(upload_props$usesProperties)
```

`getProperties()` is also available for deployments that expose
`/metadata/properties/<database>`, but rollout timing may vary by API deployment.

## 3) Search and Inspect a Category

```{r search-inspect}
if (run_live) {
  hits <- searchDatabase(
    database = "SocioMap",
    domain = "ADM0",
    term = "Ghana",
    property = "Name"
  )
  details <- CMIDinfo(database = "SocioMap", cmid = "SM1")
} else {
  hits <- list(
    count = 1,
    data = data.frame(
      CMID = "SM1",
      CMName = "Ghana",
      domain = "ADM0",
      stringsAsFactors = FALSE
    )
  )
  details <- list(node = list(CMID = "SM1", CMName = "Ghana", domain = "ADM0"))
}

hits
str(details)
```

## 4) Prepare External Input

```{r input-table}
raw_df <- data.frame(
  country_label = c("Ghana", "Cote dIvoire", "Tanzania"),
  year = c(2019, 2019, 2019),
  indicator_value = c(10.2, 9.7, 8.4),
  stringsAsFactors = FALSE
)

raw_df
```

## 5) Translate Labels

```{r translate}
if (run_live) {
  translated <- translate(
    rows = raw_df,
    database = "SocioMap",
    domain = "ADM0",
    term = "country_label",
    property = "Name",
    query = "false"
  )
} else {
  translated <- list(
    file = data.frame(
      country_label = c("Ghana", "Cote dIvoire", "Tanzania"),
      generated_CMID = c("SM1", "SM2", "SM3"),
      generated_CMName = c("Ghana", "Cote d'Ivoire", "Tanzania"),
      score = c(1, 1, 1),
      stringsAsFactors = FALSE
    )
  )
}

translated$file
```

## 6) Post-Translation Checks

```{r checks}
translated_df <- translated$file

if (!is.data.frame(translated_df) || !"generated_CMID" %in% names(translated_df)) {
  message("No `generated_CMID` column found in translation output; verify API output shape before QA checks.")
} else {
  # Rows without a CMID assignment
  translated_df[is.na(translated_df$generated_CMID), , drop = FALSE]

  # Simple duplicate check by assigned CMID
  sort(table(translated_df$generated_CMID), decreasing = TRUE)
}
```

## 7) Join Harmonized Tables

```{r join}
if (run_live) {
  left_df <- data.frame(
    datasetID = "SD1",
    country = c("Ghana", "Tanzania"),
    GID = c("GHA", "TZA"),
    metric_left = c(10, 20),
    stringsAsFactors = FALSE
  )

  right_df <- data.frame(
    datasetID = "SD2",
    country = c("Ghana", "Tanzania"),
    geonameid = c("2300660", "149590"),
    metric_right = c(2, 5),
    stringsAsFactors = FALSE
  )

  joined <- joinDatasets(
    database = "SocioMap",
    joinLeft = left_df,
    joinRight = right_df,
    domain = "CATEGORY"
  )
} else {
  joined <- data.frame(
    generated_CMID = c("SM1", "SM3"),
    left_metric = c(10, 20),
    right_metric = c(2, 5),
    stringsAsFactors = FALSE
  )
}

joined
```

## Optional: Authenticated Upload Flow

Write operations require a valid API key tied to a registered CatMapper
account and are shown as a template only. The server identifies the acting
user from the API key and enforces permissions for write endpoints.
CatMapR does not manage username/password login flows.

Key format depends on upload mode:

- `so = "standard"` (recommended) expects full key expressions in the selected key column (for example `VARIABLE == VALUE`).
- `so = "simple"` is only supported with `ao = "add_uses"` and expects raw values only (for example `eth:yoruba`) without `==`.

If preformatted key expressions are provided in `so = "simple"`, CatMapR
raises an error and requires `so = "standard"`.

After upload submission, CatMapR automatically triggers contextual USES relationship refresh in the graph database based on USES properties that connect to other CMIDs. This trigger is fire-and-forget and is not polled for completion.

```{r write-flow, eval = FALSE}
upload_payload <- data.frame(
  CMName = "Yoruba",
  Name = "Yoruba",
  CMID = "",
  Key = "Type == Adamana Brown",
  datasetID = "SD1",
  label = "ETHNICITY",
  stringsAsFactors = FALSE
)

result <- submitEditUpload(
  df = upload_payload,
  database = "SocioMap",
  formData = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    alternateCategoryNamesColumns = character(0),
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  so = "standard",
  ao = "add_uses",
  api_key = Sys.getenv("CATMAPR_API_KEY"),
  poll_interval_seconds = 1,
  timeout_seconds = 600
)

head(result)
```

```{r write-flow-update-add, eval = FALSE}
update_add_payload <- data.frame(
  CMID = "SM123",
  Key = "Type == Adamana Brown",
  datasetID = "SD1",
  variable = "CeramicType",
  stringsAsFactors = FALSE
)

submitEditUpload(
  df = update_add_payload,
  database = "SocioMap",
  formData = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  so = "standard",
  ao = "update_add",
  optionalProperties = c("variable"),
  api_key = Sys.getenv("CATMAPR_API_KEY")
)
```
