---
title: "CatMapR Safe Upload Patterns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CatMapR Safe Upload Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why this vignette

CatMapR upload flows are usually asynchronous and can mutate production
metadata. This vignette focuses on safe defaults and repeatable patterns.

## Upload mode matrix

- `so = "standard"` (recommended): provide fully formed `Key` expressions such as `Type == Adamana Brown`.
- `so = "simple"`: only valid with `ao = "add_uses"` and raw key values without `==`.

## Key helper patterns

```{r key-helpers, eval=FALSE}
build_key("Type", "Adamana Brown")
normalize_key("Key == Region == Flagstaff")
is_normalized_key("Region == Flagstaff")
```

## Add USES ties safely

```{r add-uses, eval=FALSE}
payload <- data.frame(
  CMName = "Yoruba",
  Name = "Yoruba",
  CMID = "",
  Key = "Type == Adamana Brown",
  datasetID = "SD1",
  label = "ETHNICITY",
  stringsAsFactors = FALSE
)

result <- submitEditUpload(
  df = payload,
  database = "SocioMap",
  formData = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  so = "standard",
  ao = "add_uses",
  api_key = Sys.getenv("CATMAPR_API_KEY")
)

head(result)
```

## Update existing USES properties

```{r update-add, eval=FALSE}
update_add_payload <- data.frame(
  CMID = "SM123",
  Key = "Type == Adamana Brown",
  datasetID = "SD1",
  variable = "CeramicType",
  stringsAsFactors = FALSE
)

submitEditUpload(
  df = update_add_payload,
  database = "SocioMap",
  formData = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  so = "standard",
  ao = "update_add",
  optionalProperties = c("variable"),
  api_key = Sys.getenv("CATMAPR_API_KEY")
)
```

## Replace USES key values

```{r update-replace, eval=FALSE}
update_replace_payload <- data.frame(
  CMID = "SM123",
  Key = "Type == Adamana Brown",
  NewKey = "Type == Tusayan Gray",
  datasetID = "SD1",
  stringsAsFactors = FALSE
)

submitEditUpload(
  df = update_replace_payload,
  database = "SocioMap",
  formData = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  so = "standard",
  ao = "update_replace",
  optionalProperties = c("NewKey"),
  api_key = Sys.getenv("CATMAPR_API_KEY")
)
```
