---
title: "CatMapR Safe Upload Patterns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CatMapR Safe Upload Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why this vignette

CatMapR upload flows are usually asynchronous and can mutate production
metadata. This vignette focuses on safe defaults and repeatable patterns.

## Key requirements

- Upload calls use standard key expressions.
- Provide fully formed `Key` values such as `Type == Adamana Brown`.

## Full `action` to Edit GUI mapping

This crosswalk maps CatMapR `action` values to the exact labels shown in the
CatMapper Edit page (**Advanced** upload options).

| CatMapR `action` value | Edit page GUI value | Typical purpose |
| --- | --- | --- |
| `add_node` | `Adding new node for every row` | Create new category or dataset nodes from upload rows. |
| `node_add` | `Updating existing Node properties--add or add to properties` | Add one or more properties to existing nodes without replacing current values. |
| `node_replace` | `Updating existing Node properties--replace one property` | Replace one property value on existing nodes. |
| `add_uses` | `Adding new uses ties (with old or new nodes)` | Create dataset-to-category `USES` ties. |
| `update_add` | `Updating existing USES only--add or add to properties` | Add one or more properties to existing `USES` ties. |
| `update_replace` | `Updating existing USES only--replace one property` | Replace an existing property value on `USES` ties. |
| `add_merging` | `Adding new merging ties for every row` | Create new merging ties from upload rows. |
| `merging_add` | `Updating existing Merging tie properties--add or add to properties` | Add one or more properties to existing merging ties. |
| `merging_replace` | `Updating existing Merging tie properties--replace one property` | Replace one property value on existing merging ties. |

Notes:

- Required columns vary by `action` and are validated before upload.

## Key helper patterns

```{r key-helpers, eval=FALSE}
build_key("Type", "Adamana Brown")
build_key_from_columns(
  data.frame(Type = "Adamana Brown", Region = "Flagstaff", stringsAsFactors = FALSE),
  c("Type", "Region")
)
normalize_key("Key == Region == Flagstaff")
is_normalized_key("Region == Flagstaff")
```

## Add USES ties safely

```{r add-uses, eval=FALSE}
payload <- data.frame(
  CMName = "Yoruba",
  Name = "Yoruba",
  CMID = "",
  Key = "Type == Adamana Brown",
  datasetID = "SD1",
  label = "ETHNICITY",
  stringsAsFactors = FALSE
)

result <- upload_rows(
  df = payload,
  database = "SocioMap",
  form_data = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  action = "add_uses",
  api_key = Sys.getenv("CATMAPR_API_KEY")
)

head(result)
```

## Update existing USES properties

```{r update-add, eval=FALSE}
update_add_payload <- data.frame(
  CMID = "SM123",
  Key = "Type == Adamana Brown",
  datasetID = "SD1",
  variable = "CeramicType",
  stringsAsFactors = FALSE
)

upload_rows(
  df = update_add_payload,
  database = "SocioMap",
  form_data = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  action = "update_add",
  properties = c("variable"),
  api_key = Sys.getenv("CATMAPR_API_KEY")
)
```

## Replace USES key values

```{r update-replace, eval=FALSE}
update_replace_payload <- data.frame(
  CMID = "SM123",
  Key = "Type == Adamana Brown",
  NewKey = "Type == Tusayan Gray",
  datasetID = "SD1",
  stringsAsFactors = FALSE
)

upload_rows(
  df = update_replace_payload,
  database = "SocioMap",
  form_data = list(
    domain = "ETHNICITY",
    subdomain = "ETHNICITY",
    datasetID = "SD1",
    cmNameColumn = "CMName",
    categoryNamesColumn = "Name",
    cmidColumn = "CMID",
    keyColumn = "Key"
  ),
  action = "update_replace",
  properties = c("NewKey"),
  api_key = Sys.getenv("CATMAPR_API_KEY")
)
```
